---
title: "FP_Homeenv_v1"
author: "Somendra"
date: "December 8, 2018"
output: html_document
---

```{r setup ,  message=FALSE}

library(ggbeeswarm)
library(cluster)
library(tidyverse)
library(ggthemes)
library(carData)
library(extracat)
library(pgmm)
library(GGally)
library(vcd)

student_por <- read_csv('student-por.csv')
student_mat <- read_csv('student-mat.csv')


theme_set(c(theme_classic(12),plot.title = element_text(face = "bold", size = 12) 
            ,plot.subtitle = element_text(face = "bold", color = "grey35", size = 11) 
            ,plot.caption = element_text(color = "grey68",size=5)
            ,axis.text = element_text(size=10)))
```

##### Reference to the paper
[USING DATA MINING TO PREDICT SECONDARY SCHOOL STUDENT PERFORMANCE](http://www3.dsi.uminho.pt/pcortez/student.pdf)

##### Student file contains the following number of observations : 
* Mathematics : 395
+ Portuguese  : 677
+ Both        : 382


##### Although some students have performed poorly in mathematics and have managed a passing score in portuguese, two variables are positively correlated
```{r fig0.0, fig.height = 6,fig.width = 10}

student_both=merge(student_mat,student_por,by=c("school","sex","age","address","famsize","Pstatus","Medu",
                                                "Fedu","Mjob","Fjob","reason","nursery","internet"))
#print(nrow(student_both)) # 382 students

ggplot(student_both, aes(G3.x,G3.y,color = "blue"))+
  geom_jitter() + 
  theme(legend.position = "none") +
  xlab("Maths grade") +
  ylab("Portuguese grade") +
  ggtitle("Correlation between grades in Maths & Portuguese")

```


##### Distribution of grades
```{r fig0.1, fig.height = 6,fig.width = 10}

p_por <- student_por %>% mutate(grade = as.factor(G3)) %>%
group_by(grade) %>% summarise(freq = n()) %>%
ggplot() +
geom_histogram(aes(x = grade ,y = freq), stat = 'identity') +
ggtitle("Portuguese : Frequency of students")+
xlab("Final Grade") + ylab("Number of students")
p_mat <- student_mat %>% mutate(grade = as.factor(G3))%>%
group_by(grade) %>% summarise(freq = n()) %>%
ggplot() +
geom_histogram(aes(x = grade ,y = freq), stat = 'identity') +
ggtitle("Mathematics : Frequency of students")+
xlab("Final Grade") + ylab("Number of students")
gridExtra::grid.arrange(p_por,p_mat,ncol=2)

```


##### Coverting GPA - G3, G2, G1 to erasmus system discribed in the paper
```{r, echo = FALSE}

student_por = 
student_por %>%
  mutate(G3_erasmus = if_else(between(G3, 1,9),'Fail',
                      if_else(between(G3, 10,11),'Sufficient',
                      if_else(between(G3, 12,13),'Satisfactory',
                      if_else(between(G3, 14,15),'Good','Excellent')))))

student_por$G3_erasmus <- factor(student_por$G3_erasmus, ordered = TRUE, levels =  c('Fail', 'Sufficient', 'Satisfactory', 'Good', 'Excellent'))

student_por = 
student_por %>%
  mutate(G2_erasmus = if_else(between(G2, 1,9),'Fail',
                      if_else(between(G2, 10,11),'Sufficient',
                      if_else(between(G2, 12,13),'Satisfactory',
                      if_else(between(G2, 14,15),'Good','Excellent')))))
student_por$G2_erasmus <- factor(student_por$G2_erasmus, ordered = TRUE, levels =  c('Fail', 'Sufficient', 'Satisfactory', 'Good', 'Excellent'))

student_por = 
student_por %>%
  mutate(G1_erasmus = if_else(between(G1, 1,9),'Fail',
                      if_else(between(G1, 10,11),'Sufficient',
                      if_else(between(G1, 12,13),'Satisfactory',
                      if_else(between(G1, 14,15),'Good','Excellent')))))
student_por$G1_erasmus <- factor(student_por$G1_erasmus, ordered = TRUE, levels =  c('Fail', 'Sufficient', 'Satisfactory', 'Good', 'Excellent'))

```

#### 1) Urban VS. Rural setting

##### How do the GPAs vary?

```{r fig2, fig.height = 6,fig.width = 10}

student_por$address[student_por$address == 'U'] <- "Urban"
student_por$address[student_por$address == 'R'] <- "Rural"


address_gpa<- student_por %>%
  ggplot(aes(x = address, y = G3, color = address)) +
  geom_boxplot() +
  scale_color_manual(values = c("#e15759","#a0cbe8"))+
ggtitle("Higher GPAs for student who live in Urban areas")


freq <- ggplot(student_por, aes(x = address, fill = address)) +
  geom_bar() +
  scale_fill_manual(values = c("#e15759","#a0cbe8"))+
  ggtitle("More students from urban than rural in the report")

gridExtra::grid.arrange(address_gpa,freq,ncol=2)

```

##### Internet : Do they have better access to internet?


```{r fig3, fig.height = 6,fig.width = 10}
student_por %>%
  ggplot(aes(x = internet, y = G3, color = internet)) +
  geom_boxplot() +
  facet_wrap(~address)+
  scale_color_manual(values = c("#e15759","#a0cbe8"))+
ggtitle("Higher GPAs for student who live in Urban areas")

```

##### Schools : is a particular school more linient ? and urban people go to that school?

```{r fig4, fig.height = 6,fig.width = 10}
student_por %>%
  ggplot(aes(x = school, y = G3, color = school)) +
  geom_boxplot() +
  #facet_wrap(~)+
  scale_color_manual(values = c("#e15759","#a0cbe8"))+
ggtitle("Higher GPAs for student who live in Urban areas")

```

##### Percentage urban VS schools they go to 


```{r fig5, fig.height = 6,fig.width = 10}

student_por %>% 
  group_by(school, address) %>% tally() %>%
  group_by(address) %>%
  mutate(percent = n/sum(n))  %>%
  arrange(desc(school)) %>%
  
  ggplot(aes(x = address, y = percent)) + 
  
  geom_col(aes(fill = school)) +
  geom_text(aes(label = paste(round(percent,2) * 100, "%",sep = "")),position = position_stack(vjust = 0.5))+
  scale_colour_manual(values = c("#e15759","#a0cbe8")) +
  scale_fill_manual(values = c("#e15759","#a0cbe8")) +
  labs(fill = "school") +
  ggtitle("76% urban people study at GP vs 40% rural")

```


#### 2) What role does famsize play?

```{r fig6, fig.height = 6,fig.width = 10}
famsize_gpa<- student_por %>%
  ggplot(aes(x = famsize, y = G3, color = famsize)) +
  geom_boxplot() +
  scale_color_manual(values = c("#e15759","#a0cbe8"))+
ggtitle("No difference in grades")


freq <- ggplot(student_por, aes(x = famsize, fill = famsize)) +
  geom_bar() +
  scale_fill_manual(values = c("#e15759","#a0cbe8"))+
  ggtitle("Most of the families have sizes greater than 3")

gridExtra::grid.arrange(famsize_gpa,freq,ncol=2)

```


##### What if the setting is Urban/Rural 

```{r fig7, fig.height = 6,fig.width = 10}
student_por %>%
  ggplot(aes(x = address, y = G3, color = famsize)) +
  geom_boxplot() +
  facet_wrap(~Pstatus) + 
  scale_color_manual(values = c("#e15759","#a0cbe8"))+
ggtitle("No difference in grades")

```


##### What if the parents are leaving separately?
```{r fig8, fig.height = 6,fig.width = 10}

student_por %>%
  ggplot(aes(x = famsize, y = G3, color = famsize)) +
  geom_boxplot() +
  facet_wrap(~Pstatus) + 
  scale_color_manual(values = c("#e15759","#a0cbe8"))+
ggtitle("No difference in grades")

```



##### Does parents living separately have any impact?

```{r fig9, fig.height = 6,fig.width = 10}

student_por %>%
  ggplot(aes(x = Pstatus, y = G3, color = Pstatus)) +
  geom_boxplot() +
  scale_color_manual(values = c("#e15759","#a0cbe8"))+
ggtitle("No difference in grades")


```


##### How about gender and pstatus?

```{r fighi,fig.height =0.1,fig.width = 0.1 }
x <- student_por %>%
  ggplot(aes(x = Pstatus, y = G3, color = Pstatus)) +
  geom_boxplot() +
  facet_wrap(~sex)+
  scale_color_manual(values = c("#e15759","#a0cbe8"))+
ggtitle("Males students whose parents were separated performed better")


freq_fem <- ggplot(filter(student_por,sex=="F"), aes(x = Pstatus, fill = Pstatus)) +
  geom_bar() +
  scale_fill_manual(values = c("#e15759","#a0cbe8"))+
  ggtitle("Females: parents staying together")

freq_male <- ggplot(filter(student_por,sex=="M"), aes(x = Pstatus, fill = Pstatus)) +
  geom_bar() +
  scale_fill_manual(values = c("#e15759","#a0cbe8"))+
  ggtitle("Males: parents staying together")

y <- gridExtra::grid.arrange(freq_fem,freq_male, ncol = 2)


```

```{r fig10, fig.height = 10,fig.width = 10}
gridExtra::grid.arrange(x,y,ncol=1,nrow=2)
```




